---
title: 'MAT 8444 Final Project'
author: "Katherine M. Prioli"
date: "May 08, 2019"
output:
  pdf_document: default
  html_document: default
geometry: margin = 0.5in
references:
- URL: https://fred.stlouisfed.org/series/A229RX0Q048SBEA 
  author:
  - family: Federal Reserve Bank of St. Louis
  id: fred_disposable
  issued:
    year: 2018
  title: Real Disposable Personal Income - Per Capita
- URL: https://fred.stlouisfed.org/series/TDSP
  author:
  - family: Federal Reserve Bank of St. Louis
  id: fred_debt
  issued:
    year: 2018
  title: Household Debt Service Payments as a Percent of Disposable Personal Income
- URL: https://fred.stlouisfed.org/series/A072RC1Q156SBEA
  author:
   - family: Federal Reserve Bank of St. Louis
  id: fred_savings
  issued:
    year: 2018
  title: Personal Saving as a Percent of Disposable Personal Income
- URL: https://github.com/kmprioliPROF/MAT_8444_Final_Project
  author:
  - family: Prioli
    given: Katherine M.
  id: kmpgit
  issued:
    month: 05
    year: 2019
  title: MAT_8444_Final_Project
abstract: "BACKGROUND text METHODS text RESULTS text CONCLUSION text"
---

```{r setup, include = FALSE, warning = FALSE, results = "hide", echo = FALSE, message = FALSE}
# Loading libraries

library(tidyverse)
library(forecast)     # For tidy time series analysis and graphics
library(ggthemr)      # For prettifying output
library(gridExtra)    # For grid.arrange()
library(grid)         # For arrangeGrob()
library(astsa)        # For sarima() and sarima.for()
library(scales)       # For dollar_format()

ggthemr("flat")
```


## **Background**


THe objective of this study was to analyze US household debt service payments and personal savings as they relate to real per-capita disposable income over time.

## **Methods**

Text

The analytic dataset consisted of three variables (Table 1).  All variables were obtained from the Federal Reserve Bank of St. Louis, all share a common year of valuation ($US 2012), and all have already been seasonally adjusted.


| **Variable**      	| **Description**                                                                                 	| **Reference**    	|
|-------------------	|-------------------------------------------------------------------------------------------------	|------------------	|
| `fred_disposable` 	| Per-capita disposable income, adjusted for inflation in chained $US 2012, seasonally adjusted   	| @fred_disposable 	|
| `fred_debt`       	| Household debt service payments as a percent of disposable personal income, seasonally adjusted 	| @fred_debt       	|
| `fred_savings`    	| Personal saving as a percentage of disposable personal income, seasonally adjusted              	| @fred_savings    	|

The data were be limited to a common time horizon (Q1 1980 through Q3 2018), yielding 155 observations for each time series.

### *Data Exploration*

Text


## **Results**

Text

### *Data Exploration*

```{r wrangle, echo = FALSE, message = FALSE, warning = FALSE, fig.width = 10, fig.height = 8}
fred_disposable <- read_csv("data/fred_disposable.csv") %>%
  rename(disposable = A229RX0Q048SBEA) %>%
  filter(DATE > "1979-12-31") %>% 
  select(-DATE) %>% 
  ts(start = 1980, frequency = 4)

fred_debt <- read_csv("data/fred_debt.csv") %>%
  rename(debt = TDSP) %>%
  select(-DATE) %>% 
  ts(start = 1980, frequency = 4)

fred_savings <- read_csv("data/fred_savings.csv") %>%
  rename(savings = A072RC1Q156SBEA) %>%
  filter(DATE > "1979-12-31") %>% 
  select(-DATE) %>% 
  ts(start = 1980, frequency = 4)
```

```{r exploration, echo = FALSE, message = FALSE, warning = FALSE, fig.width = 10, fig.height = 8}
fred_disp_plot <- ggtsdisplay(fred_disposable, points = FALSE,
  main = "Figure 1.  Real Disposable Personal Income Per Capita")

fred_debt_plot <- ggtsdisplay(fred_debt, points = FALSE,
  main = "Figure 2.  Household Debt Service Payments as a Percent of Disposable Personal Income")

fred_savings_plot <- ggtsdisplay(fred_savings, points = FALSE,
  main = "Figure 3.  Personal Savings as a Percent of Disposable Personal Income")
```

A generally increasing trend was seen in the plot of disposable income.  Because these data are adjusted for inflation to $US 2012, this trend represents a true increase in disposable income, not inflation, and thus the time series is not stationary and will require differencing.  The ACF plot is suggestive of an AR(p) process.  The graphs of debt payments and savings do not exhibit any simple trend, but do fluctuate over time, though without any apparent seasonality (which is expected, because the data are provided in seasonally adjusted form).  Additionally, the ACF plots for debt payments and savings are suggestive of AR(p) processes due to their slow decline.  ARCH/GARCH behavior is also possible for each of these three TS, but the ACF plots for each of the squared time series (not shown here in the interest of brevity) did not meaningfully differ from the ACF plots of the raw TS.


## **Discussion**

Text

## **Conclusion**

Moar text




Goals for the project include:

* Exploring these data through ACF, PACF
* Fitting an appropriate model to these data
* Determining what other variable(s) may be required if a regression is attempted

I'm particularly interested in understanding the decrease in savings percentage since the 1980s.  I suspect this is in part due to the Reagan-era deregulations of mortgage down payments, which disincentivized saving the traditional 20%.  Another contributing factor could be lifestyle inflation.  I also aim to forecast per-capita personal income through Q1 2021 (10 lags in the future from end of data) with a 95% prediction interval, as well as point estimates and 95% PIs for percent debt and percent savings through Q1 2021, with translation of these percentages into dollar values (point estimates and 95% PI) based on the predicted per-capita personal income.


## **References**